\texttt{dispRity} demo for ecologists

### 2015-11-26 doc/dispRity-ecology-demo.Rnw
Ecology demo update

To keep it short, this package allows to use all the dimensions of ordinated matrices (i.e. PCA, MDS, PCO) for statistical analysis rather than just a sub-set of dimensions.
For example, one might want to know whether the some sort of water treatment alters invertebrate communities and composition in natural habitats.

@@

\subsection{Data}
For this example with ecological data we are going to use data from McClean (unpubl.) that contains an ordination of a distance matrix between different experimental plots.

@@

## let's use the matrix from McClean (unpubl.)
data(McClean_data)
## This dataset contains an ordinated matrix (in 20 dimensions) of the distance
## between experimental plots.
ord_matrix <- McClean_data$ordination
## As well as two list of different factors affecting each experimental plot:
## the treatment and the depth.
treatments <- McClean_data$treatment
depth <- McClean_data$depth
@

\section{A classical two dimensional approach}
A classical way to represent this ordinated data will be to use PCA plots.

<<PCAplot, fig.width=8, fig.height=8, out.width='.8\\linewidth'>>=
## The x and y axis represent the two first dimensions of the PCA
x <- ord_matrix[, 1]
y <- ord_matrix[, 2]
## The colors will represent the treatments
cols <- sub("a", "red", treatments)
cols <- sub("b", "blue", cols)
## The symbols will represent the depth
pchs <- sub(1, 16, depth)
pchs <- as.numeric(sub(2, 17, pchs))
## Graphical option
par(bty = "n")
## A classic PCA plot
plot(x, y, col = cols, pch = pchs, xlab = "PCA1", ylab = "PCA2",
    xlim = range(x) + c(0, 100))
@

However, a common problem in this type of multivariate analysis is that often only a handful of dimensions are studied (usually the ones that bears the most variance of the ordinated matrix).

This shows the distribution of the experimental plots along the two first axis of variance of the ordinated distance matrix.
However, the problem, is it ignores the 18 other axis of the ordination and the PCA axis 1 and 2 do not represent a biological reality \textit{per se} but more some ordinations of correlations between the data and some factors.
Therefore, one might want to approach this problem without getting stuck in only two dimensions and consider the whole dataset as a \textit{n}-dimensional object.
In practice, we might be interested in looking how some experimental treatment for example, will affect the position of our experimental plots in this \textit{n}-dimensional object.
For example: do the experimental plots shift in some specific space of the \textit{n}-dimensional object when depth increases?

@@

As a first split, one might want to split the data (i.e. the \textit{n}-dimensional object) into subsamples that we want to compare.
First let's make a factor table:

<<factors>>=
## Making the factor table
factors <- as.data.frame(matrix(data = c(treatments, depth), nrow = nrow(ord_matrix),
    ncol = 2, byrow = FALSE, dimnames = list(rownames(ord_matrix))))
names(factors)<-c("Treat", "Depth")
## And here is what it looks like
head(factors)
@

Second, let's split the data according to these factors to create the subsamples of the ordinated space by using the \texttt{cust.series} function:

<<cust.series>>=
## Splitting the ordinated space into four subsamples
customised_series <- cust.series(ord_matrix, factors)
## Note that the output of dispRity functions are dispRity objects
class(customised_series)
## These objects are automatically printed in a summary method (calling S3 print.dispRity)
## giving information about the object
customised_series
@

For more details on the \dispRity objects, see the \href{https://github.com/TGuillerme/dispRity}{other dispRity demos}.
Basically the idea is to avoid jamming the \R console such as when using:

<<dispRity class, eval=FALSE>>=
## Summarise the object
str(customised_series)
@

@@

Now we're going to see the functionalities of the core function of this package: the \dispRity function.
This function is a modulable function that allow to simply (and quickly!) calculate disparity from a matrix.
Disparity can be calculated in many ways, this function is a tool to measure disparity \textit{as defined by the user} (and here's where the modulable part comes in).
For more details on disparity, see the \href{https://github.com/TGuillerme/dispRity}{other dispRity demos}.

One can usually decompose the disparity metrics into two elements:
\begin{enumerate}
\item the \textbf{class metric} that is a descriptor of the matrix. For example describing the ranges of each column in the matrix or the euclidean distances between each row and the centroid of the matrix.
\item the \textbf{summary metric} that is a summary of the class metric values. For example, the sum of the ranges or the median of the euclidean distances. 
\end{enumerate}
Basically the combination can be infinite between the class and summary metrics.
For example, people might want to measure the median variances of the axis or the product of the distances from the centroid.
However, it is probable that some metrics are better to reflect some biological aspects of the any-o-space than others...

In practice, the \dispRity function intakes a pair of class and summary metrics as a definition of disparity.
Several of these metrics are implemented in other packages (like \texttt{stats::median}, \texttt{base::sum}, etc.) and this package proposes several metrics listed in \texttt{dispRity.metric} (see \texttt{?dispRity.metric}).
But it is even possible to use your very own class and summary metrics! Or even any metric you decide makes sense (note that this is not yet implemented in v.\Sexpr{packageVersion("dispRity")}).
This will be actually heavily encouraged and facilitate with the \texttt{make.metric} function in a future version.

To use these metrics pairs in the \dispRity function, it's pretty easy:

<<dispRity1>>=
## For example, let's calculate the median distance between each plot and the
## centroid of the ordinated space
disparity <- dispRity(customised_series, metric = c(median, centroids))
## Note that disparity is a dispRity object and printing it just gives details
## on the object, not the results. We need to use summary.dispRity (S3) to get
## the results.
summary(disparity)
@

Note that we calculated the median distance between plots and the centroid but the output displays mean values.
This is because summary will, by default, summarise the data using the mean value.
Here the mean represents the mean of the median distance for each series which is a bit useless (i.e. the mean of one value is that same value).
We can display the median as well using:

<<dispRity2>>=
summary(disparity, cent.tend=median)
## Or even the product! It won't affect the results
summary(disparity, cent.tend=prod)
@

This is because we did calculate the noise within our data.
We can classically do so by bootstrapping the data!

\subsection{Bootstrapping the data}

The \dispRity pacakge also provides easy way to bootstrap the data via the \texttt{boot.matrix} function.
We can even rarefy the data to see the effect of the number of plots per series:
For more details on the bootstrapping options, see the \href{https://github.com/TGuillerme/dispRity}{other dispRity demos}.

<<boot.matrix>>=
bootstrapped_data <- boot.matrix(customised_series, bootstraps=100)
rarefied_data <- boot.matrix(customised_series, bootstraps=100, rarefaction=TRUE)
## Note that the output is a dispRity object giving some details on the series and the bootstraps
bootstrapped_data
@

We can now rerun a more robust disparity analysis:

<<dispRity3>>=
disparity_BS <- dispRity(bootstrapped_data, metric = c(median, centroids))
disparity_rare <- dispRity(rarefied_data, metric = c(median, centroids))
## Note that calculation time is increased!
@

@@

<<summary>>=
## The default:
summary(disparity_BS)
## The quantiles are calculated as 50 and 95 and the central tendency is the mean by default.
## But we can specify different options
summary(disparity_BS, quantile=90, cent.tend=median)

## Finally we can see the results of the rarefaction analysis:
head(summary(disparity_rare))
## This outputs a longer table with all the variations of plots down to 3 plots per series.
@

Finally we can also plot the results using the simple \texttt{plot.dispRity S3} method:

<<disparity plots, fig.width=8, fig.height=8, out.width='.8\\linewidth'>>=
## Graphical option
par(bty = "n")
## Plotting the score for each groups
plot(disparity_BS)
@

Or have a look at the effect of the number of experimental plots:

<<rare plots, fig.width=10, fig.height=5, out.width='1\\linewidth'>>=
## Graphical options
quartz(width = 10, height = 5) ; par(mfrow = (c(1,2)), bty = "n")
## The same but looking at the number of plots
plot(disparity_rare, elements = TRUE)
## With the same number of plots per group
plot(disparity_rare, elements = TRUE, rarefaction = 17)
@

Or event have a look at the rarefaction curves:

<<rare plots2, fig.width=8, fig.height=8, out.width='.8\\linewidth'>>=
## Graphical option
par(bty = "n")
## Plotting the rarefaction curves
plot(disparity_rare, rarefaction = "plot")
@

\subsection{Testing the hypothesis}

\bibliographystyle{sysbio}
\bibliography{References}

>>>>

To keep it short, this package allows to use all the dimensions of ordinated matrices (i.e. PCA, MDS, PCO) for statistical analysis rather than just a sub-set of dimensions.
For example, one might want to know whether the sort of water treatment at certain depth alters invertebrate communities and composition in natural habitats.

@@

\subsection{Data}
For this example with ecological data we are going to use data from McClean (unpubl.). %INC CITE!
This data is the ordination of a distance matrix based on nutrient enrichment and depth and freshwater benthic invertebrates.

@@

## Loading the data
data(McClean_data)

## This dataset contains an ordinated matrix (20 dimensions) of the distance
## between 40 experimental plots.
ord_matrix <- McClean_data$ordination

## As well as two list of different factors affecting each experimental plot:
## the treatment and the depth.
treatments <- McClean_data$treatment
depth <- McClean_data$depth
@

\section{A classical two dimensional approach}
A classical way to represent this ordinated data will be to use two dimensional plots.

<<PCAplot, fig.width=8, fig.height=8, out.width='.8\\linewidth'>>=
## The x and y axis represent the two first dimensions of the ordination
x <- ord_matrix[, 1]
y <- ord_matrix[, 2]
## The colors will represent the treatments
cols <- sub("a", "red", treatments)
cols <- sub("b", "blue", cols)
## The symbols will represent the depth
pchs <- sub(1, 16, depth)
pchs <- as.numeric(sub(2, 17, pchs))

## Graphical option
par(bty = "n")
## A classic 2D ordination plot
plot(x, y, col = cols, pch = pchs, xlab = "Dimension 1", ylab = "Dimension 2",
    xlim = range(x) + c(0, 100))
@

This shows the distribution of the experimental plots along the two first axis of variance of the ordinated distance matrix (i.e. the first two dimensions).
However, the problem, is that this plot ignores the 18 other dimensions of the ordination!
Additionally, these two represented dimensions do not represent a biological reality \textit{per se}; i.e. the values on the first dimension do not represent some continuous traits (e.g. depth) but just the ordinations of correlations between the data and some factors.

Therefore, one might want to approach this problem without getting stuck in only two dimensions and consider the whole dataset as a \textit{n}-dimensional object.
To explore our hypothesis, we could look if there are differences in the hyper-volume that different treatments occupy in the ordinated space to see the effect of water treatments on the communities.

@@

The first step in such analysis will be to create different series that are two sub-samples of the ordinated space (i.e. the \textit{n}-dimensional object).
Each of these series contain a certain number of elements (i.e. the 40 experimental field plots) that have some attributes in common.
In our example, we are going to group the elements according to their depth and treatment.

<<factors>>=
## Creating the table that contain the elements and their attributes
factors <- as.data.frame(matrix(data = c(treatments, depth), nrow = nrow(ord_matrix),
    ncol = 2, byrow = FALSE, dimnames = list(rownames(ord_matrix))))
names(factors)<-c("Treat", "Depth")
head(factors)
@

Second, let's split the data according to these factors to create the series of the ordinated space by using the \texttt{cust.series} function:

<<cust.series>>=
## Splitting the ordinated space into four subsamples
customised_series <- cust.series(ord_matrix, factors)

## Note that the output of dispRity functions are dispRity objects
class(customised_series)
## These objects are automatically printed in a summary method (calling S3 print.dispRity)
## giving information about the object
customised_series
@

For more details on the \dispRity objects, see the \href{https://github.com/TGuillerme/dispRity}{\dispRity manual}.
Basically the idea is to avoid jamming the \R console such as when using:

<<dispRity class, eval=FALSE>>=
## Summarising the object
str(customised_series)

## Displaying the full object
print.dispRity(customised_series, all=TRUE)
@

@@

Disparity can be calculated in many ways, and therefore, the \dispRity function allows to measure disparity \textit{as defined by the user}.
For more details on disparity, see the \href{https://github.com/TGuillerme/dispRity}{other vignettes}.

In this example, we are going to define disparity as the median distance between the different experimental plots and the centroid of the ordinated space.
High values of disparity will indicate a general high dispersal from this centroid (i.e. on average, the experimental plots are far apart in the ordinated space).
We can define the metrics easily in the \dispRity function by feeding them to the \texttt{metric} argument.
Here we are going to feed the functions \texttt{median::stats} and \texttt{centroids::dispRity} which calculates distances between elements and centroid.

<<dispRity1>>=
## Calculating disparity
disparity <- dispRity(customised_series, metric = c(median, centroids))

## Note that disparity is a dispRity object and printing it just gives details
## on the object, not the results. We need to use summary.dispRity (S3) to get
## the results.
disparity
summary(disparity)
@

\subsection{Bootstrapping the data}
One might also want to bootstrap the data to test the robustness of the measured disparity to outliers.
Also, as we can see, each series has different numbers of elements.
It might be interesting to rarefy the data as well to have only series with the same number of elements.
Both steps are easily doable through the \texttt{boot.matrix} function.

<<boot.matrix>>=
## Simple bootstrapping (1000 times)
bootstrapped_data <- boot.matrix(customised_series, bootstraps = 100)
## Bootstrapping with rarefaction (i.e. only re-sampling 17 elements each time)
rarefied_data <- boot.matrix(customised_series, bootstraps = 100, rarefaction = 17)
## Note that the output is a dispRity object giving some details on the series and the bootstraps
bootstrapped_data
rarefied_data
@

\noindent We can now rerun a more robust disparity analysis using the bootstrapped data:

<<dispRity3>>=
## Bootstrapped disparity
disparity_BS <- dispRity(bootstrapped_data, metric = c(median, centroids))
## Rarefied disparity
disparity_rare <- dispRity(rarefied_data, metric = c(median, centroids))
@

@@

<<summary>>=
## Summarising disparity (default)
summary(disparity_BS)
## The quantiles are calculated as 50 and 95 and the central tendency is the mean by default.
## But we can specify different options
summary(disparity_BS, quantile=90, cent.tend=median)

## Finally we can see the results of the rarefaction analysis:
summary(disparity_rare)
## Note that for the first three series, there is no observed disparity since the data was
## rarefied for these series.
@

\noindent We can also plot the results and have a look at the effect of the number of experimental plots:

<<rare plots, fig.width=10, fig.height=5, out.width='1\\linewidth'>>=
## Graphical options
quartz(width = 10, height = 5) ; par(mfrow = (c(1,2)), bty = "n")
## Plotting the bootstrapped disparity
plot(disparity_BS, main="Bootstrapped data")
## Plotting the rarefied disparity
plot(disparity_rare, main="Rarefied data")
@

\noindent As we can see, there seems to be no strong effect of the number of experimental plots in each series (i.e. the rarefied plot is really similar to the bootstrapped plot) which is a good thing!

\subsection{Testing the hypothesis}
Finally, we can test our hypothesis (whether the sort of water treatment at certain depth alters invertebrate communities and composition in natural habitats) by using the \texttt{test.dispRity} function.

<<test>>=
## Testing the effect of our factors on the bootstrapped data
summary(test.dispRity(disparity_BS, test = aov, comparisons = "all"))
## Testing the effect of our factors on the rarefied data
summary(test.dispRity(disparity_rare, test = aov, comparisons = "all"))

## Post-hoc testing of the effect of the two different factors
## Note that the comparison contains the list of the pairs of series to compare
test.dispRity(disparity_BS, test = wilcox.test, comparison = list(c(1,2), c(3,4)))
## Testing the effect of the two different factors for the rarefied data
test.dispRity(disparity_rare, test = wilcox.test, comparison = list(c(1,2), c(3,4)))
@

\noindent As we can see, there is strong support for an effect of the treatment and the depth on the median distance between each experimental plots and the centroid of the ordinated space.
In other words, it seems that with the second treatment and the second level of depth, the invertebrate communities and composition were further apart (i.e. more dispersed in the ordinated space) than with the first treatment and the first level of depth.

% \bibliographystyle{sysbio}
% \bibliography{References}
